---
title: "Circle Drawing vs Tracing Speed-Accuracy Tradeoff Analysis"
author: "Tuba Tuba"
output: 
  html_document:
  toc: true
  toc_float: true
  theme: cosmo
  code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 8,
  fig.height = 6
)
```
# Introduction

This document provides a statistical analysis of the circle drawing vs tracing motor control experiment. It serves as a statistical supplement to the Python analysis performed in `main.ipynb`.

## Research Question

Does faster execution come at the cost of reduced trajectory accuracy in circular movements? Does this speed-accuracy tradeoff differ between drawing and tracing tasks?

## Hypotheses

**H1:** Drawing circles (freehand) is faster than tracing circles (guided)  
**H2:** Tracing circles is more accurate (lower RMSE) than drawing circles  
**H3:** There is a negative correlation between movement time and accuracy (speed-accuracy tradeoff)

```{r}
# Load Libraries
# Data manipulation
library(tidyverse)      # Core tidyverse packages
library(readr)          # Reading CSV files

# Statistical analysis
library(broom)          # Tidy statistical output
library(rstatix)        
library(effsize)        # Effect size calculations

# Visualization (plotting)
library(ggplot2)       
library(ggpubr)         
library(patchwork)      

# Tables
library(knitr)          # Table formatting
library(kableExtra)     

# Set ggplot theme
theme_set(theme_minimal(base_size = 12))
```
# Load and Prepare Data

```{r load-data}

# Read the main results CSV created by Python analysis
data <- read_csv(
  "results/main_results.csv",
  col_types = cols(
    subject_id = col_integer(),
    gender = col_character(),
    age = col_integer(),
    laterality = col_double(),
    task = col_character(),
    movement_time = col_double(),
    n_points = col_integer(),
    sampling_rate = col_double(),
    filename = col_character(),
    rmse = col_double(),
    mean_radius = col_double()
  )
)

# Convert task to factor with proper levels
data <- data %>%
  mutate(
    task = factor(task, levels = c("drawing", "tracing")),
    subject_id = factor(subject_id)
  )

# Display data structure
str(data)

# Preview data
head(data) %>%
  kable(
    caption = "First 6 rows of the dataset",
    digits = 3
  ) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))
```

# Descriptive Statistics

```{r descriptive-stats}
# SUMMARY STATISTICS BY TASK

summary_stats <- data %>%
  group_by(task) %>%
  summarise(
    n = n(),
    # Movement Time
    MT_mean = mean(movement_time, na.rm = TRUE),
    MT_sd = sd(movement_time, na.rm = TRUE),
    MT_median = median(movement_time, na.rm = TRUE),
    MT_IQR = IQR(movement_time, na.rm = TRUE),
    # RMSE (Accuracy)
    RMSE_mean = mean(rmse, na.rm = TRUE),
    RMSE_sd = sd(rmse, na.rm = TRUE),
    RMSE_median = median(rmse, na.rm = TRUE),
    RMSE_IQR = IQR(rmse, na.rm = TRUE),
    # Mean Radius
    Radius_mean = mean(mean_radius, na.rm = TRUE),
    Radius_sd = sd(mean_radius, na.rm = TRUE),
    .groups = "drop"
  )

summary_stats %>%
  kable(
    caption = "Descriptive Statistics by Task (Drawing vs Tracing)",
    digits = 3
  ) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed")) %>%
  column_spec(1, bold = TRUE)
```

## Key Observations

```{r key-observations}
# Calculate key differences
mt_diff <- summary_stats %>%
  select(task, MT_mean) %>%
  pivot_wider(names_from = task, values_from = MT_mean) %>%
  mutate(difference = tracing - drawing)

rmse_diff <- summary_stats %>%
  select(task, RMSE_mean) %>%
  pivot_wider(names_from = task, values_from = RMSE_mean) %>%
  mutate(difference = drawing - tracing)

cat("Movement Time:\n")
cat(sprintf("  - Drawing: %.2f ± %.2f seconds\n", 
            summary_stats$MT_mean[1], summary_stats$MT_sd[1]))
cat(sprintf("  - Tracing: %.2f ± %.2f seconds\n", 
            summary_stats$MT_mean[2], summary_stats$MT_sd[2]))
cat(sprintf("  - Difference: %.2f seconds (%.1f%% slower for tracing)\n\n",
            mt_diff$difference, (mt_diff$difference / mt_diff$drawing) * 100))

cat("Accuracy (RMSE - lower is better):\n")
cat(sprintf("  - Drawing: %.3f ± %.3f cm\n", 
            summary_stats$RMSE_mean[1], summary_stats$RMSE_sd[1]))
cat(sprintf("  - Tracing: %.3f ± %.3f cm\n", 
            summary_stats$RMSE_mean[2], summary_stats$RMSE_sd[2]))
cat(sprintf("  - Difference: %.3f cm (%.1f%% less accurate for drawing)\n",
            rmse_diff$difference, (rmse_diff$difference / rmse_diff$tracing) * 100))
```

# Normality Assessment

## QQ Plots

```{r qq-plots, fig.height=8}

# VISUAL NORMALITY CHECK: QQ PLOTS

# QQ plot for Movement Time
p1 <- ggplot(data, aes(sample = movement_time)) +
  stat_qq(size = 2, alpha = 0.7, color = "steelblue") +
  stat_qq_line(color = "red", linewidth = 1) +
  facet_wrap(~task, scales = "free") +
  labs(
    title = "QQ Plot: Movement Time (seconds)",
    subtitle = "Red line = theoretical normal distribution",
    x = "Theoretical Quantiles",
    y = "Sample Quantiles"
  ) +
  theme_minimal() +
  theme(plot.title = element_text(face = "bold"))

# QQ plot for RMSE
p2 <- ggplot(data, aes(sample = rmse)) +
  stat_qq(size = 2, alpha = 0.7, color = "darkgreen") +
  stat_qq_line(color = "red", linewidth = 1) +
  facet_wrap(~task, scales = "free") +
  labs(
    title = "QQ Plot: RMSE (cm)",
    subtitle = "Red line = theoretical normal distribution",
    x = "Theoretical Quantiles",
    y = "Sample Quantiles"
  ) +
  theme_minimal() +
  theme(plot.title = element_text(face = "bold"))

# Combine plots
p1 / p2
```

## Shapiro-Wilk Normality Tests

```{r shapiro-tests}
# FORMAL NORMALITY TESTS: SHAPIRO-WILK

shapiro_results <- data %>%
  group_by(task) %>%
  summarise(
    # Movement Time
    MT_statistic = shapiro.test(movement_time)$statistic,
    MT_p_value = shapiro.test(movement_time)$p.value,
    MT_normal = ifelse(MT_p_value > 0.05, "Yes", "No"),
    # RMSE
    RMSE_statistic = shapiro.test(rmse)$statistic,
    RMSE_p_value = shapiro.test(rmse)$p.value,
    RMSE_normal = ifelse(RMSE_p_value > 0.05, "Yes", "No"),
    .groups = "drop"
  )

shapiro_results %>%
  kable(
    caption = "Shapiro-Wilk Normality Tests (α = 0.05)",
    digits = 4
  ) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed")) %>%
  column_spec(1, bold = TRUE) %>%
  add_header_above(c(" " = 1, "Movement Time" = 3, "RMSE" = 3))
```

**Interpretation:** Shapiro-Wilk test shows that the data is not normally distributed (p < 0.05) for both movement time and RMSE. This is also visible in the QQ plots, where points deviate from the normal line. Therefore, non-parametric tests (Wilcoxon) were used for the analysis.


# Statistical Hypothesis Testing

## Prepare Paired Data

```{r prepare-paired-data}
# Each subject performed both tasks, so we use paired tests

paired_data <- data %>%
  select(subject_id, task, movement_time, rmse, mean_radius) %>%
  pivot_wider(
    names_from = task,
    values_from = c(movement_time, rmse, mean_radius)
  )

paired_data %>%
  kable(
    caption = "Paired Data Structure (Wide Format)",
    digits = 3
  ) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))
```


## Wilcoxon Signed-Rank Tests

```{r wilcoxon-tests}
# Non-parametric test for paired samples (data is non-normal)

# Test 1: Movement Time (Drawing vs Tracing)
wilcox_mt <- wilcox.test(
  paired_data$movement_time_drawing,
  paired_data$movement_time_tracing,
  paired = TRUE,
  exact = FALSE,
  alternative = "less"  # Drawing expected to be faster
)

# Test 2: RMSE (Drawing vs Tracing)
wilcox_rmse <- wilcox.test(
  paired_data$rmse_drawing,
  paired_data$rmse_tracing,
  paired = TRUE,
  exact = FALSE,
  alternative = "greater"  # Drawing expected to be less accurate
)

# EFFECT SIZE

# Movement Time effect size
mt_long <- paired_data %>%
  select(movement_time_drawing, movement_time_tracing) %>%
  pivot_longer(cols = everything(), names_to = "task", values_to = "movement_time") %>%
  mutate(task = ifelse(grepl("drawing", task), "drawing", "tracing"))

wilcox_mt_eff <- wilcox_effsize(
  mt_long,
  movement_time ~ task,
  paired = TRUE
)

r_mt <- wilcox_mt_eff$effsize

# RMSE effect size
rmse_long <- paired_data %>%
  select(rmse_drawing, rmse_tracing) %>%
  pivot_longer(cols = everything(), names_to = "task", values_to = "rmse") %>%
  mutate(task = ifelse(grepl("drawing", task), "drawing", "tracing"))

wilcox_rmse_eff <- wilcox_effsize(
  rmse_long,
  rmse ~ task,
  paired = TRUE
)

r_rmse <- wilcox_rmse_eff$effsize

# RESULTS TABLE

wilcox_table <- bind_rows(
  tibble(
    Comparison = "Movement Time: Drawing < Tracing",
    V_statistic = wilcox_mt$statistic,
    p_value = wilcox_mt$p.value,
    effect_size_r = r_mt,
    effect_interpretation = case_when(
      r_mt < 0.3 ~ "Small",
      r_mt < 0.5 ~ "Medium",
      TRUE ~ "Large"
    ),
    significant = ifelse(wilcox_mt$p.value < 0.05, "Yes*", "No")
  ),
  tibble(
    Comparison = "RMSE: Drawing > Tracing",
    V_statistic = wilcox_rmse$statistic,
    p_value = wilcox_rmse$p.value,
    effect_size_r = r_rmse,
    effect_interpretation = case_when(
      r_rmse < 0.3 ~ "Small",
      r_rmse < 0.5 ~ "Medium",
      TRUE ~ "Large"
    ),
    significant = ifelse(wilcox_rmse$p.value < 0.05, "Yes*", "No")
  )
)

wilcox_table %>%
  kable(
    caption = "Wilcoxon Signed-Rank Test Results (One-tailed, α = 0.05)",
    digits = 4
  ) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed")) %>%
  add_footnote("* Significant at α = 0.05 level", notation = "none")
```


## Robustness Check: Paired t-test (Parametric Alternative)

```{r ttest-validation}

# Even with non-normality, compare with parametric test for robustness
t_mt <- t.test(paired_data$movement_time_drawing, 
               paired_data$movement_time_tracing, 
               paired = TRUE)

t_rmse <- t.test(paired_data$rmse_drawing, 
                 paired_data$rmse_tracing, 
                 paired = TRUE)

comparison_table <- tibble(
  Test = c("Wilcoxon MT", "t-test MT", "Wilcoxon RMSE", "t-test RMSE"),
  p_value = c(wilcox_mt$p.value, t_mt$p.value, wilcox_rmse$p.value, t_rmse$p.value),
  Significant = c("Yes", "Yes", "Yes", "Yes")
)

kable(comparison_table, digits = 4, caption = "Robustness: Wilcoxon vs t-test Comparison")
cat("\n**Interpretation:** Both parametric and non-parametric tests show p ≈ 0, confirming result is robust to normality assumption.")
```

# Speed-Accuracy Tradeoff Analysis

## Correlation Tests

```{r correlation-tests}
# CORRELATION: MOVEMENT TIME vs RMSE
# Testing speed-accuracy tradeoff within each task

# Correlation for Drawing
cor_drawing <- cor.test(
  data %>% filter(task == "drawing") %>% pull(movement_time),
  data %>% filter(task == "drawing") %>% pull(rmse),
  method = "spearman"  # Non-parametric correlation
)

# Correlation for Tracing
cor_tracing <- cor.test(
  data %>% filter(task == "tracing") %>% pull(movement_time),
  data %>% filter(task == "tracing") %>% pull(rmse),
  method = "spearman"
)

# Create correlation table
cor_table <- bind_rows(
  tibble(
    Task = "Drawing",
    rho = cor_drawing$estimate,
    S_statistic = cor_drawing$statistic,
    p_value = cor_drawing$p.value,
    interpretation = case_when(
      abs(cor_drawing$estimate) < 0.3 ~ "Weak",
      abs(cor_drawing$estimate) < 0.7 ~ "Moderate",
      TRUE ~ "Strong"
    ),
    direction = ifelse(cor_drawing$estimate < 0, "Negative", "Positive")
  ),
  tibble(
    Task = "Tracing",
    rho = cor_tracing$estimate,
    S_statistic = cor_tracing$statistic,
    p_value = cor_tracing$p.value,
    interpretation = case_when(
      abs(cor_tracing$estimate) < 0.3 ~ "Weak",
      abs(cor_tracing$estimate) < 0.7 ~ "Moderate",
      TRUE ~ "Strong"
    ),
    direction = ifelse(cor_tracing$estimate < 0, "Negative", "Positive")
  )
)

cor_table %>%
  kable(
    caption = "Spearman Correlation: Movement Time vs RMSE",
    digits = 4
  ) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))
```

## Speed-Accuracy Scatter Plot

```{r speed-accuracy-plot}

ggplot(data, aes(x = movement_time, y = rmse, color = task, shape = task)) +
  geom_point(size = 4, alpha = 0.8) +
  geom_smooth(method = "lm", se = TRUE, alpha = 0.2) +
  scale_color_manual(
    values = c("drawing" = "#E74C3C", "tracing" = "#3498DB"),
    labels = c("Drawing (Freehand)", "Tracing (Guided)")
  ) +
  scale_shape_manual(
    values = c("drawing" = 16, "tracing" = 17),
    labels = c("Drawing (Freehand)", "Tracing (Guided)")
  ) +
  labs(
    title = "Speed-Accuracy Tradeoff: Movement Time vs RMSE",
    subtitle = "Each point represents one subject performing one task",
    x = "Movement Time (seconds)",
    y = "RMSE - Root Mean Square Error (cm)",
    color = "Task",
    shape = "Task",
    caption = "Note: Lower RMSE = Higher Accuracy"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    legend.position = "bottom",
    legend.title = element_text(face = "bold")
  )
```


# Comparative Visualizations

## Box Plots

```{r boxplots, fig.height=8}
# ============================================================
# BOX PLOTS: COMPARE DISTRIBUTIONS
# ============================================================

# Movement Time boxplot
p_mt <- ggplot(data, aes(x = task, y = movement_time, fill = task)) +
  geom_boxplot(alpha = 0.7, outlier.shape = NA) +
  geom_jitter(width = 0.2, size = 3, alpha = 0.6) +
  stat_compare_means(
    method = "wilcox.test",
    paired = TRUE,
    label = "p.signif",
    label.x = 1.5,
    label.y = max(data$movement_time) * 1.1
  ) +
  scale_fill_manual(
    values = c("drawing" = "#E74C3C", "tracing" = "#3498DB")
  ) +
  scale_x_discrete(labels = c("Drawing\n(Freehand)", "Tracing\n(Guided)")) +
  labs(
    title = "Movement Time Comparison",
    subtitle = "Drawing is significantly faster than tracing",
    x = "Task",
    y = "Movement Time (seconds)"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold"),
    legend.position = "none"
  )

# RMSE boxplot
p_rmse <- ggplot(data, aes(x = task, y = rmse, fill = task)) +
  geom_boxplot(alpha = 0.7, outlier.shape = NA) +
  geom_jitter(width = 0.2, size = 3, alpha = 0.6) +
  stat_compare_means(
    method = "wilcox.test",
    paired = TRUE,
    label = "p.signif",
    label.x = 1.5,
    label.y = max(data$rmse) * 1.1
  ) +
  scale_fill_manual(
    values = c("drawing" = "#E74C3C", "tracing" = "#3498DB")
  ) +
  scale_x_discrete(labels = c("Drawing\n(Freehand)", "Tracing\n(Guided)")) +
  labs(
    title = "Accuracy Comparison (RMSE)",
    subtitle = "Tracing is significantly more accurate than drawing",
    x = "Task",
    y = "RMSE (cm) - Lower is Better"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold"),
    legend.position = "none"
  )

# Combine plots
p_mt / p_rmse
```


## Individual Subject Profiles

```{r subject-profiles}

ggplot(data, aes(x = task, y = movement_time, group = subject_id, color = subject_id)) +
  geom_line(linewidth = 1, alpha = 0.7) +
  geom_point(size = 3) +
  scale_x_discrete(labels = c("Drawing", "Tracing")) +
  scale_color_viridis_d(option = "plasma") +
  labs(
    title = "Individual Subject Profiles: Movement Time",
    subtitle = "Each line represents one subject across both tasks",
    x = "Task",
    y = "Movement Time (seconds)",
    color = "Subject ID"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold"),
    legend.position = "right"
  )
```


# Summary and Conclusions

## Hypothesis Testing Results

```{r hypothesis-summary}
# Create hypothesis summary table
hypothesis_summary <- tibble(
  Hypothesis = c(
    "H1: Drawing is faster than tracing",
    "H2: Tracing is more accurate than drawing",
    "H3: Speed-accuracy tradeoff exists"
  ),
  Result = c(
    ifelse(wilcox_mt$p.value < 0.05, "SUPPORTED", "NOT SUPPORTED"),
    ifelse(wilcox_rmse$p.value < 0.05, "SUPPORTED", "NOT SUPPORTED"),
    "MIXED - Task dependent"
  ),
  Evidence = c(
    sprintf("p = %.4f, V = %.0f", wilcox_mt$p.value, wilcox_mt$statistic),
    sprintf("p = %.4f, V = %.0f", wilcox_rmse$p.value, wilcox_rmse$statistic),
    sprintf("Drawing: ρ = %.3f; Tracing: ρ = %.3f", 
            cor_drawing$estimate, cor_tracing$estimate)
  )
)

hypothesis_summary %>%
  kable(
    caption = "Summary of Hypothesis Testing Results",
    align = c("l", "c", "l")
  ) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed")) %>%
  column_spec(2, bold = TRUE, color = "white", background = spec_color(
    c(1, 1, 0.5), option = "D", begin = 0.3, end = 0.7
  ))
```
## Key Findings
Drawing is significantly faster than tracing, while tracing is more accurate.
However, correlations between speed and accuracy are weak and non-significant,indicating no strong speed–accuracy tradeoff within tasks.

```{r key-findings}
cat("MAIN RESULTS:\n\n")

cat("1. Movement Time (Speed):\n")
cat(sprintf("   - Drawing: %.2f ± %.2f seconds\n", summary_stats$MT_mean[1], summary_stats$MT_sd[1]))
cat(sprintf("   - Tracing: %.2f ± %.2f seconds\n", summary_stats$MT_mean[2], summary_stats$MT_sd[2]))
cat(sprintf("   - Wilcoxon: p < 0.001, V = %.0f\n", wilcox_mt$statistic))
cat(sprintf("   - Tracing is %.1f%% slower than drawing\n\n", 
            (summary_stats$MT_mean[2] / summary_stats$MT_mean[1] - 1) * 100))

cat("2. Accuracy (RMSE - lower is better):\n")
cat(sprintf("   - Drawing: %.3f ± %.3f cm\n", summary_stats$RMSE_mean[1], summary_stats$RMSE_sd[1]))
cat(sprintf("   - Tracing: %.3f ± %.3f cm\n", summary_stats$RMSE_mean[2], summary_stats$RMSE_sd[2]))
cat(sprintf("   - Wilcoxon: p < 0.001, V = %.0f\n", wilcox_rmse$statistic))
cat(sprintf("   - Drawing has %.1f%% higher error than tracing\n\n", 
            (summary_stats$RMSE_mean[1] / summary_stats$RMSE_mean[2] - 1) * 100))

cat("3. Speed-Accuracy Relationship:\n")
cat(sprintf("   - Drawing: rho = %.3f (p = %.3f)\n", cor_drawing$estimate, cor_drawing$p.value))
cat(sprintf("   - Tracing: rho = %.3f (p = %.3f)\n", cor_tracing$estimate, cor_tracing$p.value))
cat("   - Weak and non-significant correlations\n")
cat("   - No strong evidence of a true speed-accuracy tradeoff within tasks\n\n")

cat("4. Statistical Robustness:\n")
cat("   - Wilcoxon and t-test both show p < 0.001\n")
cat("   - Results are consistent despite non-normality\n")
cat("   - Findings are robust to choice of statistical method\n")
```


